---
title: "What factors affect fuel effiency of a car?"
author: "DomR"
date: "Saturday, September 13, 2014"
output: pdf_document
---

## Executive Summary
This report analyis data from 32 1973-1974 model cars and tries to explore if there are other factors other than transmission type that can be used to predict fuel efficiency of a car.

* This report specially tries to answer the following questions:
    + Is an automatic or manual transmission better for MPG?
    + Quantify the MPG difference between automatic and manual transmissions
    
    
## Data Preparation
```{r, echo=TRUE,message=FALSE}
require(MASS)
require(plyr)
require(ggplot2)
require(lattice)
require(knitr)
options("scipen"=100, "digits"=4)
opts_chunk$set(fig.width=7, fig.height=5, tidy=FALSE, size='small',width=100 )
```

### Load the data

```{r, echo=TRUE}
data(mtcars)
#Convert factor variables into factors
mtcars$cyl <- as.factor(mtcars$cyl) #4, 6 or 8
mtcars$vs <- as.factor(mtcars$vs) #0- V engine or 1= straight engine. 
mtcars$am <- as.factor(mtcars$am) #0 - automatic and 1 - gear
mtcars$gear <- as.factor(mtcars$gear) #3,4,5
mtcars$carb <- as.factor(mtcars$carb) #1,2,3,4
```

## Exploratory Data Analysis

### Base model using am as the only predictor variable keeping all other factors constant
```{r, echo=TRUE}
baseFit <- lm(mpg~am, data=mtcars)
summary(baseFit)
autoSummary <- summary(subset(mtcars, am==0)$mpg)
manualSummary <- summary(subset(mtcars, am==1)$mpg)
```
* The mean, median, minimum and maximum mpg (see also Figure 1 in Appendix) for 
    + automatic cars  is `r autoSummary["Mean"]`,`r autoSummary["Median"]`,`r autoSummary["Min."]` and `r autoSummary["Max."]` respectively.
    + manual cars is `r manualSummary["Mean"]`,`r manualSummary["Median"]`,`r manualSummary["Min."]` and `r manualSummary["Max."]` respectively.

*  The manual tranmission ( am1) cars provide additional **7.245** miles per gallon of gas compared to automatic cars. However, adjusted R-squared value is only `r summary(baseFit)$adj.r.squared`. _Is there a better model than the baseFit?_

### Identify the best model to predict miles per gallon for a car.
```{r, echo=TRUE, results="hide"}
#Get a lm fit of mpg against all other factors
lmFit <- lm(mpg~., data=mtcars)
#Perform a step wise model selection of mpg versus other factors
steplmFit <- stepAIC(lmFit, direction="both")
steplmFit$anova
```
```{r, echo=TRUE}
attr(terms(steplmFit),"term.label")
```
* As shown, the significant predictor variables to predict mpg outcome is `r attr(terms(steplmFit),"term.label")` and the best Fit model is
```{r, echo=TRUE}
bestFitWithCylHpWtAm <- lm(mpg~cyl + hp + wt + am, data=mtcars)
summary(bestFitWithCylHpWtAm)
```
*  Adjusted R-square for best fit model is `r summary(bestFitWithCylHpWtAm)$adj.r.squared` which is much better than base model. How much transmission type has effect on mpg? Let's build multiple models with cyl, hp and wt as predictor variables.

### How much does transmission type really impacts mpg?
####  Build few models model without using am as predictor variable

```{r, echo=TRUE}
bestFitWithCylHpWt <- lm(mpg~cyl + hp + wt, data=mtcars)
#summary(bestFitWitCylHpWt)
bestFitWithCylWt <- lm(mpg~cyl + wt, data=mtcars)
#summary(bestFitWithCylWt)
bestFitWithWt <- lm(mpg~wt, data=mtcars)
#summary(bestFitWithWt)
bestFitWithCyl <- lm(mpg~cyl, data=mtcars)
#summary(bestFitWithCyl)

rSquaredValues <- rbind(
c("Model : mpg~cyl + hp + wt + am",summary(bestFitWithCylHpWtAm)$adj.r.squared),
c("Model : mpg~cyl + hp + wt ",summary(bestFitWithCylHpWt)$adj.r.squared),
c("Model : mpg~cyl + wt",summary(bestFitWithCylWt)$adj.r.squared),
c("Model : mpg~wt",summary(bestFitWithWt)$adj.r.squared),
c("Model : mpg~cyl ",summary(bestFitWithCyl)$adj.r.squared)
)
colnames(rSquaredValues ) <-c("Model", "Adj. R-Squared")
```
*  Adjusted R-square for the new models 
```{r, echo=TRUE}
rSquaredValues
```
Let's further analyze these models using anova.

#### Compare models using pValues
```{r, echo=TRUE}
modelCompare <- anova(bestFitWithCylHpWtAm, bestFitWithCylHpWt,bestFitWithCylWt,bestFitWithWt,bestFitWithCyl)
modelCompare
```
  + As seen above from pValues, model differences between _bestFitWithCylHpWtAm_ and _bestFitWitCylHpWt_ is  not significant ( pValue of 0.206 > 0.05 (95% confidence level), Null hypothesis of models been equal cannot be rejected). Model differences between _bestFitWithCylHpWt_ and _bestFitWithCylWt_ is also not significant( pValue of 0.06 > 0.05). However, model differences bestFitWithCylWt and bestFitWithWt are significant (pValue of 0.001733 > 0.05 and hence Null hypothesis of models been same can be rejected). Also, rValue of model _bestFitWithCylWt_ (.82) is much higher than that of _bestFitWithWt_ (0.74).  Since we prefer a model with least number of predictor variables, we can conclude model _bestFitWithCylWt_ (with cylinder and weigth) is the best model for predicting fuel efficiency rather than tramission type.

### Quantity mpg differences between automatic and manual cars

Let's add _am_ to the best model _bestFitWithCylWt_
```{r, echo=TRUE}
bestFitWithCylWtAm <- lm(mpg~cyl + wt+am, data=mtcars)
summary(bestFitWithCylWtAm)
```
As seen from above, the manual cars get  just **0.150** additional miles per gallon when number of cylinders and weight of a car are taken into consideration.

### Residual Analysis
#### Residual analysis for model _bestFitWithCylWt_ using wt and cyl as predictor variables

```{r, echo=TRUE}
bestFitWithCylWt$residuals
```
Toyoto Corolla (5.79 ), Fiat 128 (5.46), Chrysler Imperial(3.91), Toyoto Corona (-4.58) ( see also Figure 2 in the Appendix) are the outliers in the dataset with either very high/low efficieny which will have an effect of using model _bestFitWithCylWt_ for prediction.


## Conclusion

Using _baseFit_ with only transmission type as the predictor variable shows that manual tranmission type cars achieve higher efficiency of  **7.2** miles per gallon than automatic cars. However, as shown above, transmission type is a not a good predictor for fuel efficiency, but rather weight and number of cylinders of a car. Taking number of cylinders and weight into consideration, manual cars provide just **0.15** additional miles per gallon than automatic cars.


---

## Appendix

### Figure 1 showing a boxplot comparing mpg for automatic and manual transmission cars
```{r, echo=TRUE}
with(mtcars,{
     boxplot(mpg ~ am, 
             ylab = "miles per gallon",
             xlab = "0 - auto, 1 - manual",
             main = "Fig. 1 - mpg for auto and manual tramission cars")
    abline(baseFit, col="red")
     })
```

### Figure 2 showing diagnostic plots for best model ( mpg ~ cyl  + wt).

```{r, echo=TRUE}
par(mfrow = c(2,2))
plot(bestFitWithCylWt)
```